Community-Conditioned Language Models

### 2020-11-24 paper/paper.tex
ppl = e ^ (avg H), not avg(e^H)
move big table to appendix

\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.16}

@@

In the present work, we further investigate linguistic variation
across online communities. In contrast to previous work, we
investigate variation across speech communities (rather than
macro-social categories), and use context-sensitive sequence models
capable of capturing variation beyond the lexical level.
%
We do so by investigate two questions.  First, we are interested in
variation in the level of linguistic complexity exhibited by different
communities.
\jp{
  Some examples (perhaps include if there is space):

@@

We measure the perplexity of each example\jp{message? community?} in the test set. 
For this, we use the negative exponential of the mean cross-entropy loss 
of the language model for a given message and average over a set of messages, $M$.
\[\Ppl_M = \avg_{m\in M}(e^{-H(m)})\]
\bn{Following the logic from IG, should we change Ppl to the exponential of the average cross entropy, rather than using the average of the message-level perplexity?}
\jp{Yes}
\[\Ppl_M = e^{\avg_{m\in M} H(m)}\]
As one would expect, 
the conditioned models have lower perplexity
than their respective unconditioned baseline models (\cref{tab:model-results}),
but this effect is different in magnitude
for models with different community embedding depths.
For the LSTM encoder, 
the best model ($l_c=1$) concatenates the community embedding after the first encoder layer,
with two subsequent LSTM layers. 
The best transformer model ($l_c=3$), on the other hand,
incorporates the community information last,
after all the sequence encoder layers.

\begin{table}
  \footnotesize
  \centering
  \input{floats/model_results.tex}
  \caption{
    Performance of baseline (first row for each encoder architecture) and CCLM models. 
    Perplexity and information gain are shown for the entire test set 
    (\num{5000} messages for each community).}
  \label{tab:model-results}
\end{table}

We also consider model performance stratified by community; that is,
$\Ppl_{M_j}$, where $M_j$ is the test-set messages from community
$c_j$.  We observe a great deal of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{53.07} to
\num{115.29} for the LSTM model (\cref{tab:community-results}).

@@

The conditioned models also perform differently across different communities---%
for some communities, $\IG_{M_j}$ around \num{1}, or even slightly lower,
meaning that the CCLM has no advantage over the unconditioned baseline for that community.
For others communities it is higher, meaning that the CCLM performs better than 
the baseline for those communities (\cref{tab:community-results}).

\begin{table*}
  \centering
  \input{floats/community_results.tex}
  \caption{
    Model perplexity and information gain by community. We show the absolute perplexity 
    of the baseline models and the information gain of the CCLMs.
    Rows are sorted by the perplexity of the baseline
    LSTM model.
}
  \label{tab:community-results}
\end{table*}

@@

\begin{figure*}
\begin{tikzpicture}
  \begin{groupplot}[
    group style={group size=2 by 1},
    yticklabels={,,}, xticklabels={,,},
    legend style={anchor=north, legend columns=1, font=\sffamily\tiny},
  ]
    \nextgroupplot
    \commscatter{lstm-3-1_pca1}{lstm-3-1_pca2}
    \nextgroupplot
    \commscatter{transformer-3-3_pca1}{transformer-3-3_pca2}
    \legend{General interest,Videogames,Technology,Sports,Female-focused,Other}
  \end{groupplot}
\end{tikzpicture}
\caption{Community embedding PCA of the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:comm-pca}
\end{figure*}

\begin{table}
  \centering
  \input{floats/comm_sim.tex}
  \caption{Pearson's $r$ between the pairwise similarity of community vectors
  in the CCLM models and the membership-based embedding of \citet{Kumar2018}.
  ($p<0.001$ for all models).
  }
  \label{tab:pairwise-comm-sim}
\end{table}

@@

We can characterize the quality the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal dominates the matrix: the LMCC generally finds the right community for a message (\cref{fig:confusion}).
%
\begin{figure*}
\begin{tikzpicture}
  \confusiongroupplot{lstm-3-1}{transformer-3-3}
\end{tikzpicture}
\caption{LMCC confusion matrices for the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:confusion}
\end{figure*}
%
But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (46 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTM
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.98$ and $p < 0.0001$ (see, for example,
\cref{fig:lmcc-ppl}).

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
    \commscatter{lstm-3-1_indisc}{transformer-3-3_indisc}
    \end{axis}
  \end{tikzpicture}
  \caption{%
    Correlation between indiscernibility for 
    the best LSTM ($c = 1$) and transformer ($c = 3$) models
    (Pearson's $r = 0.99$, $p < 0.0001$).
  }
  \label{fig:lmcc-ppl}
\end{figure}

@@

\begin{table}
  \centering
  \input{floats/ind_corrs.tex}
  \caption{ Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$).  }
  \label{tab:ind-corrs}
\end{table}

\begin{figure*}
  \begin{tikzpicture}
  \begin{groupplot}[
    group style={group size=2 by 1},
    legend style={anchor=north, legend columns=1, font=\tiny},
    every node near coord/.append style={font=\sffamily\tiny}%
  ]
    \nextgroupplot
    \commscatter[]{lstm-3-1_lm_ppl}{lstm-3-1_indisc}
    \nextgroupplot
    \commscatter[]{lstm-3-1_info_gain}{lstm-3-1_indisc}
    \legend{General interest,Videogames,Technology,Sports,Female-focused,Other}
  \end{groupplot}
  \end{tikzpicture}
  \caption{%
    Mean community indiscernibility (y-axis), 
    in relationship to linguistic complexity (left) and 
    CCLM information gain (right). Shown here for the LSTM
    model with $l_c=1$. See \cref{tab:ind-corrs} for 
    Pearson's r correlations for all models.
  }
  \label{fig:cclm-lmcc-ppl}
\end{figure*}

@@

Along these lines, \citet{OConnor2010} uses a Bayesian generative
model to infer communities from variation in text data.  In contrast
to our work, this model treats words as independent events, ignoring
the structure (and variation) in the construction of sequences.  It
does further suggest, however, that community-level variation could be
modeled in an unsupervised way.

@@

We find that various communities exhibit widely varying levels of
linguistic complexity \ref{hyp:varying-complexity}, with perplexity
per-word ranging from less than 53 to more than 104 for the best model
architecture.

@@

We find no statistically significant correlation (\cref{tab:ind-corrs}) between the
perplexity of the
% \bn{I'm not sure if we can make this claim or not with the Ppl-Ppl correlations. However, given that *all* of them have positive correlation, even though the p-values are somewhat higher than .1 maybe we can?}
CCLM and the indiscernibility of the corresponding community
\ref{hyp:rich-harder-to-identify}. Because we have relatively few, and
only large communities in our test set, it is difficult to say if this
finding generalizes. Regardless, this kind of correlation would suggest that
community-specific complexity can be at least partially attributed to
using more varied patterns. This is view is supported by the
observation that, with an indiscernibility $> 0.9$, the most complex
communities appear as an even mixture of the kind of language found in
all communities, to our models.

We find that the layer at which the community embedding is taken into
account in a CCLM influences its perplexity \ref{hyp:layer-effect},
albeit weakly. For LSTM models, the perplexity per word, averaged over
all communities, varies between 76.80 and 78.64. (with 80.70 for the
unconditioned model) The perplexity per word is slightly higher for
$l_c=0$, and has very little difference between the other models, with
$l_c=1$ being the best.  For transformer models, it varies between
89.38 and 122.56. Here perplexity per-word is clearly not a monotonous
function of $l_c$: the maximum is reached at $l_c=2$. We did not
expect non-monotonicity, and leave studying this effect to future
work.

>>>>

\usepgfplotslibrary{groupplots}
\usepgfplotslibrary{statistics}
\pgfplotsset{compat=1.16}
\pgfplotsset{every tick label/.append style={font=\footnotesize}}

@@

In the present work, we further investigate linguistic variation
across online communities. In contrast to previous work, we
investigate variation across speech communities (rather than
macro-social categories), and use context-sensitive sequence models
capable of capturing variation beyond the lexical level.
%
We do so by investigating two questions.  First, we are interested in
variation in the level of linguistic complexity exhibited by different
communities.
\jp{
  Some examples (perhaps include if there is space):

@@

We measure the perplexity of each message in the test set. 
For this, we use the negative exponential of the mean cross-entropy loss 
of the language model for a given message and average over a set of messages, $M$.
\[\Ppl_M = e^{\avg_{m\in M} H(m)}\]
As one would expect, 
the conditioned models have lower perplexity
than their respective unconditioned baseline models (\cref{tab:model-results}),
but this effect is different in magnitude
for models with different community embedding depths.
For the LSTM encoder, 
the best model ($l_c=1$) concatenates the community embedding after the first encoder layer,
with two subsequent LSTM layers. 
The best transformer model ($l_c=3$), on the other hand,
incorporates the community information last,
after all the sequence encoder layers.

\begin{table}
  \small
  \centering
  \input{floats/model_results.tex}
  \caption{
    Performance of baseline (first row for each encoder architecture) and CCLM models. 
    Perplexity and information gain are shown for the entire test set 
    (\num{5000} messages for each community).}
  \label{tab:model-results}
\end{table}

We also consider model performance stratified by community; that is,
$\Ppl_{M_j}$, where $M_j$ is the test-set messages from community
$c_j$.  We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{34.47} to
\num{66.43} for the LSTM model (see table 1 in the supplementary materials).

\begin{figure*}[t]
\begin{tikzpicture}
  \begin{groupplot}[
    group style={group size=2 by 1},
    yticklabels={,,}, xticklabels={,,},
    legend style={anchor=north, legend columns=1, font=\sffamily\tiny},
  ]
    \nextgroupplot
    \commscatter{lstm-3-1_pca1}{lstm-3-1_pca2}
    \nextgroupplot
    \commscatter{transformer-3-3_pca1}{transformer-3-3_pca2}
    \legend{General interest,Videogames,Technology,Sports,Female-focused,Other}
  \end{groupplot}
\end{tikzpicture}
\caption{Community embedding PCA of the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:comm-pca}
\end{figure*}

\begin{table}
  \small
  \centering
  \input{floats/comm_sim.tex}
  \caption{Pearson's $r$ between the pairwise similarity of community vectors
  in the CCLM models and the membership-based embedding of \citet{Kumar2018}.
  ($p<0.001$ for all models).
  }
  \label{tab:pairwise-comm-sim}
\end{table}

@@

The conditioned models also perform differently across different communities---%
for some communities, $\IG_{M_j}$ around \num{1}, or even slightly lower,
meaning that the CCLM has no advantage over the unconditioned baseline for that community.
For other communities it is higher, meaning that the CCLM performs better than 
the baseline for those communities (\cref{fig:comm-info-gain}).

\begin{figure}
\begin{tikzpicture}
  \infogainboxplot
\end{tikzpicture}
\caption{
  Distribution of information gain across communities for each CCLM 
  (top: LSTM models, bottom: transformer models).
}
\label{fig:comm-info-gain}
\end{figure}

This variation in information gain across communities is not 
related to perplexity, however. 
Some of the most linguistically complex communities benifit a lot
from community conditioning, while others do not benefit at all. 
Similarly, among the less linguistically complex communites 
there some have high information gain, and some have information
gain of \num{1} or less (\cref{fig:ppl-info-gain}).

\begin{figure}
\begin{tikzpicture}
\begin{axis}
\commscatter[]{lstm-3-1_lm_ppl}{lstm-3-1_info_gain}
\end{axis}
\end{tikzpicture}
\caption{
  Average community perplexity (x-axis) versus information gain (y-axis)
  for the LSTM model with $l_c = 1$ .
}
\label{fig:ppl-info-gain}
\end{figure}

@@

@@

We can characterize the quality the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal dominates the matrix: the LMCC generally finds the right community for a message (\cref{fig:confusion}).
%
\begin{figure*}
%\begin{tikzpicture}
  %\confusiongroupplot{lstm-3-1}{transformer-3-3}
%\end{tikzpicture}
\caption{LMCC confusion matrices for the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:confusion}
\end{figure*}
%
But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (46 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTM
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.98$ and $p < 0.0001$ (see, for example,
\cref{fig:indisc-corr}).

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
    \commscatter{lstm-3-1_indisc}{transformer-3-3_indisc}
    \end{axis}
  \end{tikzpicture}
  \caption{%
    Correlation between indiscernibility for 
    the best LSTM ($c = 1$) and transformer ($c = 3$) models
    (Pearson's $r = 0.99$, $p < 0.0001$).
  }
  \label{fig:indisc-corr}
\end{figure}

@@

\begin{table}
  \small
  \centering
  \input{floats/ind_corrs.tex}
  \caption{ Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$).  }
  \label{tab:ind-corrs}
\end{table}

\begin{figure*}
  \begin{tikzpicture}
  \begin{groupplot}[
    group style={group size=2 by 1},
    legend style={anchor=north, legend columns=1, font=\tiny},
    every node near coord/.append style={font=\sffamily\tiny}%
  ]
    \nextgroupplot
    \commscatter[]{lstm-3-1_lm_ppl}{lstm-3-1_indisc}
    \nextgroupplot
    \commscatter[]{lstm-3-1_info_gain}{lstm-3-1_indisc}
    \legend{General interest,Videogames,Technology,Sports,Female-focused,Other}
  \end{groupplot}
  \end{tikzpicture}
  \caption{%
    Average community indiscernibility (y-axis), 
    in relationship to linguistic complexity (left) and 
    CCLM information gain (right). Shown here for the LSTM
    model with $l_c=1$. See \cref{tab:ind-corrs} for 
    Pearson's r correlations for all models.
  }
  \label{fig:cclm-lmcc-ppl}
\end{figure*}

@@

Along these lines, \citet{OConnor2010} uses a Bayesian generative
model to infer communities from variation in text data.  In contrast
to our work, this model treats words as independent events, ignoring
the structure (and variation) in the construction of sequences.  It
does further suggest, however, that community-level variation can be
modeled in an unsupervised way.

@@

We find that various communities exhibit widely varying levels of
linguistic complexity \ref{hyp:varying-complexity}, with perplexity
per-word ranging from less than 53 to more than 104 for the best model \bn{update}
architecture.

@@

We find no statistically significant correlation (\cref{tab:ind-corrs}) between the
perplexity of the
CCLM and the indiscernibility of the corresponding community
\ref{hyp:rich-harder-to-identify}. Because we have relatively few, and
only large communities in our test set, it is difficult to say if this
finding generalizes. Regardless, this kind of correlation would suggest that
community-specific complexity can be at least partially attributed to
using more varied patterns. This is view is supported by the
observation that, with an indiscernibility $> 0.9$, the most complex
communities appear as an even mixture of the kind of language found in
all communities, to our models.

We find that the layer at which the community embedding is taken into
account in a CCLM influences its perplexity \ref{hyp:layer-effect},
albeit weakly. For LSTM models, the perplexity per word, averaged over
all communities, varies between 76.80 and 78.64. (with 80.70 for the
unconditioned model) The perplexity per word is slightly higher for
$l_c=0$, and has very little difference between the other models, with
$l_c=1$ being the best.  For transformer models, it varies between \bn{update}
89.38 and 122.56. Here perplexity per-word is clearly not a monotonous
function of $l_c$: the maximum is reached at $l_c=2$. We did not
expect non-monotonicity, and leave studying this effect to future
work.

### 2020-11-24 paper/plot_macros.tex
move big table to appendix

\newcommand{\commscatter}[3][\community]{%
  \addplot[
  scatter, only marks,
  font=\tiny,
  scatter src=explicit symbolic,
  scatter/classes={
    general-interest={mark=*, blue},
    videogames={mark=*, red},
    technology={mark=*, green},
    sports={mark=*, cyan},
    female-focused={mark=*, magenta},
    other={mark=o, black}%
  },
  nodes near coords*={#1},
  nodes near coords align={vertical},
  every node near coord/.append style={font=\sffamily\tiny},
  visualization depends on={value \thisrow{community} \as \community}
  ] table [x={#2}, y={#3}, meta=category] {floats/comm.csv};
}

@@

>>>>

\newcommand{\commscatter}[3][\community]{%
  \addplot[
  scatter, only marks,
  font=\tiny,
  scatter src=explicit symbolic,
  scatter/classes={
    general-interest={mark=*, blue},
    videogames={mark=*, red},
    technology={mark=*, green},
    sports={mark=*, cyan},
    female-focused={mark=*, magenta},
    other={mark=o, black}%
  },
  nodes near coords*={#1},
  nodes near coords align={vertical},
  every node near coord/.append style={font=\sffamily\tiny},
  visualization depends on={value \thisrow{community} \as \community},
  ] table [x={#2}, y={#3}, meta=category] {floats/comm.csv};
}

@@

\newcommand{\infogainboxplot}{
  \begin{groupplot}[
    group style={
      group size=1 by 2,
      every plot/.style={
        xmin=0.65,
        xmax=1.25,
      },
    },
    height=4cm, width=8cm,
    ytick={1,2,3,4,5,6,7,8},
    yticklabels={0, 1, 2, 3, 0, 1, 2, 3},
    ylabel={$l_c$},
    ]
  \nextgroupplot
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-0_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-1_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-2_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-3_info_gain] {floats/comm.csv};
  \nextgroupplot
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-0_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-1_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-2_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-3_info_gain] {floats/comm.csv};
  \end{groupplot}
}
